---
title: "Text Analysis"
---

```{r echo=FALSE}
#| include=FALSE
library(tidyverse)
library(tidytext)
library(textdata)
library(wordcloud)
library(wordcloud2)
library(viridis)
library(ggthemes)
library(gutenbergr)
library(tidytext)
library(igraph)
library(ggraph)
library(ggplot2)
```

```{r}
#| echo: false #“Run the code, but don’t show the code in the output.”
#| include: false #run the code but show nothing
#Load the dataset
article_dat <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-02-25/article_dat.csv')

#I want only three columns
academic_Literature <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-02-25/article_dat.csv') |> 
  select("title", "abstract", "keywords")

```

```{r}
#| warning: false
#| echo: false #“Run the code, but don’t show the code in the output.”
#| eval: false
install.packages("DT", repos = "https://cloud.r-project.org")
```

```{r}
#| echo: false #“Run the code, but don’t show the code in the output.”
#| eval: false
```

### Introduction
This project explores text analysis and sentiment analysis using a dataset of academic literature on racial and ethnic disparities in reproductive medicine in the United States. The primary goal is to develop practical skills in text mining, including the use of regular expressions, tokenization, and data visualization techniques.

The analysis begins by transforming the text into individual words (tokenization), followed by an exploration of the most frequently occurring terms. Common stopwords are then removed to focus on meaningful content. Finally, sentiment analysis is applied to classify the text as positive, negative, or neutral, providing insight into the overall tone and themes present in the literature.

### Data Frame
```{r echo = FALSE}
#| include=FALSE
#I want to remove some symbols from my data from the title variable.  
filter1 <- academic_Literature |>
  mutate(newtitle = gsub("\\[.*?\\]|[;:0-9/()]", "", title)) |> 
  filter(title != " ") |> #remove empty rows or space
  mutate(new_keyword = gsub("\\[.*?\\]|[;:0-9/()]", "", keywords)) |>
  filter(keywords != " ")
view(filter1)
```

```{r}
#| echo: false #“Run the code, but don’t show the code in the output.”
library(DT)

datatable(
  filter1 |> 
    select(New_title = newtitle, 
           New_keyword = new_keyword) |>  # pick key columns
    head(10),
  options = list(pageLength = 2))
```

### Data Exploration 

#### Which title is the longest and how mnany words are there in it? 

```{r}
#| message: false
#| warning: false
#| echo: false #“Run the code, but don’t show the code in the output.”
data88 <- filter1 |> 
  mutate(word_count = lengths(str_split(newtitle, " "))) |>
  arrange(desc(word_count)) |>
  slice_head(n = 5)

data88 |>
  mutate(newtitle = str_remove_all(newtitle, "[[:punct:]]")) |>
  select(
    NewTitle = newtitle,
    WordCount = word_count
  ) |>
  slice_head(n = 5) |>
  knitr::kable(caption = "Longest Titles")
```

#### Does the longest title always have the most characters?

```{r}
#| message: false
#| warning: false
#| echo: false #“Run the code, but don’t show the code in the output.”

data89 <- filter1 |>
  mutate(char_count = str_length(str_remove_all(newtitle, " "))) |>
  arrange(desc(char_count)) |>
  slice_head(n = 3)

#Space is excluded
data89 |>
  mutate(
    newtitle = str_remove_all(newtitle, "[[:punct:]]"),
    newtitle = str_squish(newtitle),
    char_count = str_length(str_remove_all(newtitle, " "))
  ) |>
  select(
    NewTitle = newtitle,
    CharCount = char_count
  ) |>
  knitr::kable(caption = "Longest Titles (by Character Count)")
```

No, the longest title doees not always have the most characters. 

### Tokenization

#### Tokenizing the Title Variable. 

```{r}
#| echo: false #“Run the code, but don’t show the code in the output.”
new_data1 <- filter1 |>
  mutate(line = row_number()) |>
  unnest_tokens(word, newtitle, token = "words")

new_data1 |>
  slice_head(n = 10) |>
  select(
    Line = line,
    Words = word
  ) |>
  knitr::kable(
  caption = "<strong style='font-size:18px; color:black;'>10 Tokenized Words (Title)</strong>",
  escape = FALSE)
```

#### Tokenizing the Keywords Variable

```{r}
#| echo: false #“Run the code, but don’t show the code in the output.”
new_data2 <- filter1 |>
  mutate(line = row_number()) |> #assigns index or number to each by starting from 1. 
  unnest_tokens(word, new_keyword, token = "words")

new_data2 |>
  slice_head(n = 10) |>
  select(
    Line = line,
    Words = word
  ) |>
  knitr::kable(
  caption = "<strong style='font-size:18px; color:black;'>10 Tokenized Words (Keywords)</strong>",
  escape = FALSE)

```

### Stopwords Exploration

```{r}
#| echo: false #“Run the code, but don’t show the code in the output.”
#| include: false #run the code but show nothing
 #GET STOPWORDS DIFFERENT SOURCES
snowball <- get_stopwords(source = "snowball")
smart_stopwords <- get_stopwords(source = "smart")
nltkstop <- get_stopwords(source = "nltk")
```

#### Snowball

```{r}
#| message: false
#| warning: false
#| echo: false #“Run the code, but don’t show the code in the output.”
library(DT)

datatable(
  snowball,
  options = list(
    pageLength = 5,   # show 10 rows at a time
    autoWidth = TRUE
  ),
  caption = "Snowball Stopwords"
)
```

#### Let's remove the stopwords above from the Word variable (title) and create a table. 

```{r}
#| message: false
#| warning: false
#REMOVE STOPWORDS & COUNT WORD FREQUENCY
new_data3 <- new_data1 |>
  anti_join(smart_stopwords) |>
  count(word, sort = TRUE) |> 
  filter(word != "NA") |>
  slice_max(n, n = 30)

```


```{r}
#| message: false
#| warning: false
#| echo: false #“Run the code, but don’t show the code in the output.”

new_data3 |>
  slice_head(n = 10) |>
  select(
    Words = word,
    Count = n
  ) |>
  knitr::kable(
  caption = "<strong style='font-size:18px; color:black;'>10 Most Common Words In Title After Removing Stopwords</strong>",
  escape = FALSE)
```

### Word Frequency Bar Chart

```{r}
#| echo: false #“Run the code, but don’t show the code in the output.”
#Let's see how it looks like in a graph
new_data3 |>
  ggplot(aes(fct_reorder(word, n), n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Most Frequent Words in Literature Dataset",
    x = "Words",
    y = "Frequency") + 
  theme_minimal()
```

This horizontal bar chart displays the frequency of the most common words in the dataset, where longer bars indicate higher word frequency. Words such as “cancer,” “disparities,” and “racial” appear most frequently, suggesting a strong focus on healthcare, particularly in oncology and health disparities. Because the chart is flipped, the word labels are shown on the y-axis, while the frequency values are displayed along the x-axis. This layout makes it easier to read longer words and compare their frequencies visually. Overall, the distribution highlights a clear thematic emphasis on cancer research, patient outcomes, and demographic disparities, providing insight into the primary topics present in the literature.

#### Let's Explore The Bing Sentiment Lexicons

```{r}
#| echo: false #“Run the code, but don’t show the code in the output.”
#| include: false
#Let's get the sentiment lexicon
afin1_sentiment <- get_sentiments(lexicon = "afinn") #Each word has a numeric score (e.g., −5 to +5)
nrc1_sentiment <- get_sentiments(lexicon = "nrc") #Words are labeled with emotions
```

```{r}
#| message: false
#| warning: false
#| echo: false #“Run the code, but don’t show the code in the output.”
bing1_sentiment <- get_sentiments(lexicon = "bing") #Classifies words as: positive and negative

bing1_sentiment |>
  slice_head(n = 10) |>
  select(
    Words = word,
    Sentiment = sentiment) |>
  knitr::kable(
  caption = "<strong style='font-size:18px; color:black;'>10 Example Words with Sentiments</strong>",
  escape = FALSE)
```


### Implement Sentiment Analysis with the BING

#### Are titles more positve, negative, or neutral?

```{r}
#| message: false
#| warning: false
#| echo: false #“Run the code, but don’t show the code in the output.”
#| include: FALSE

#I want to keep only words that have a match with the bing sentiment lexicons. 
new_data3 |>
  inner_join(bing1_sentiment)

#Let's check if the title is more positive or negative
new_data3 |>
  inner_join(bing1_sentiment) |>
  group_by(sentiment) %>%          # group by positive/negative
  summarise(avg_n = mean(n))       # calculate average count per sentiment
```

#### Data Table
```{r}
#| message: false
#| warning: false
#| echo: false #“Run the code, but don’t show the code in the output.”
new_data3 |>
  inner_join(bing1_sentiment) |>
  knitr::kable(caption = "Top Words with Sentiment Labels")
```

#### First: Let's visualize with bar chart

```{r}
#| message: false
#| warning: false
#| echo: false #“Run the code, but don’t show the code in the output.”
#What words contribute the most to sentiment scores for these titles? 
new_data3 |>
  inner_join(bing1_sentiment) |>
  count(sentiment, word, sort = TRUE) |>
  group_by(sentiment) |>
  slice_max(n, n = 10) |>
  ungroup() |>
  ggplot(aes(x = fct_reorder(word, n), y = n, fill = sentiment)) +
    geom_col() +
    coord_flip() +
    facet_wrap(~ sentiment, scales = "free") +
  labs(title = "Word Frequencies by Sentiment in Academic Literature",
    x = "Words",
    y = "Frequency") + 
  theme_minimal()
```

This faceted horizontal bar chart illustrates word frequencies categorized by sentiment in academic literature. Words associated with negative sentiment, including, cancer, risk, severe, appear in the left panel, while words associated with positive sentiment, survival, appear in the right panel. Bar lengths indicate relative word frequencies. The word that contributes the most to the sentiment scores for these titles is survival. All three terms—risk, cancer, and survival—show a maximum relative frequency of $1.00$ within their respective sentiment categories.

#### Second: Let's Visualize Using Wordcloud.

```{r}
#| message: false
#| warning: false
#| echo: false #“Run the code, but don’t show the code in the output.”
# wordcloud wants a column with words and another column with counts
w11 <- new_data3 |>
  anti_join(stop_words) |>
  count(word) |>
  filter(!is.na(word)) |>
  arrange(desc(n))

wordcloud(words = w11$word,
          freq = w11$n,
          max.words = 100,
          random.order = FALSE)
```

This word cloud visualizes the most frequently occurring terms in the literature dataset. The size of each word represents its frequency, making the most dominant themes immediately apparent. The largest words—including analysis, birth, black, cancer, and, risk, indicate that these are the most prominent terms in the text. Words are arranged in both horizontal and vertical orientations to create a compact thematic summary.

### N-gram Analysis with Abstracts

#### First: I will begin by breaking the abstracts into bigrams, and the table belo is just showing only 10 example bigrams. 

```{r}
#| echo: false #“Run the code, but don’t show the code in the output.”
LiteratureNgram <- academic_Literature |> 
  group_by(abstract) |> 
  mutate(linenumber = row_number()) |>
  ungroup() |>
  unnest_tokens(bigram,  abstract, token = "ngrams", n = 2) |> 
  filter(bigram != "NA")

LiteratureNgram |>
  slice_head(n = 10) |>
  select(
    Linenumber = linenumber, 
    Bigram = bigram) |>
  knitr::kable(
  caption = "<strong style='font-size:18px; color:black;'>10 Example Bigram Words</strong>",
  escape = FALSE)
```

#### Second: Check the most common bigram

```{r}
#| echo: false #“Run the code, but don’t show the code in the output.”
BB1 <- LiteratureNgram |> 
  count(bigram, sort = TRUE)

BB1 |>
  slice_head(n = 10) |>
  select(
    Bigram = bigram, 
    Count = n) |>
  knitr::kable(
  caption = "<strong style='font-size:18px; color:black;'>10 Most Common Bigram Words</strong>",
  escape = FALSE)
```

The most common bigram is “95 ci”, which occurs in a total of 502 times.

#### Third: Remove Some of the Stopwrods From the Bigram Column

```{r}
#| echo: false #“Run the code, but don’t show the code in the output.”
bigrams_filter1 <- LiteratureNgram |>
  separate(bigram, c("firstword", "secondword"), sep = " ") |>
  filter(!firstword %in% stop_words$word,
         !secondword %in% stop_words$word) |>
  filter(!is.na(firstword) & !is.na(secondword))

#After removing some of the stop-words, I reunited both columns of the bigram to be one column. 
bigrams_filter2 <- bigrams_filter1 |>
  unite(bigram, firstword, secondword, sep = " ")

#I want to remove all the digits from the bigram column before calculating the tf score. 
bigrams_filter3 <- bigrams_filter2 |>
  mutate(bigram = gsub("[0-9,.]", "", bigram)) |> #remove all the digits from the bigram
  filter(bigram != " ") |> #remove the empty columns
  mutate(title = gsub("[0-9,.]", "", title)) |> #remove all the digits from the title
  filter(title != " ") |> #remove the empty columns
  mutate(word_count = str_count(title, "\\S+")) |> #count how many words per each title
  arrange(desc(word_count)) #arrange from highest word_count to the lowest word_count

#Extra Data with Title

#Here I want to know how many title contain only 5 words, 10 words, or 20 words. 
bigrams_filter4 <- bigrams_filter3 |>
  count(word_count)

#I want to keep only one title from each type of title because there are many titles are the same.
bigrams_filter5 <- bigrams_filter3 |>
  distinct(title, .keep_all = TRUE) |>
  mutate(word_in_title = str_count(title, "\\S+")) #count how many words per each title

```


```{r}
#| message: false
#| warning: false
#| echo: false #“Run the code, but don’t show the code in the output.”
bigrams_filter3 |>
  slice_head(n = 10) |>
  select(Linenumber = linenumber,
    Bigram = bigram) |>
  knitr::kable(
  caption = "<strong style='font-size:18px; color:black;'>10 Example Bigram Words Without Stopwords</strong>",
  escape = FALSE)
```


#### Fourth: Creating a Network Graph From the Bigram

```{r}
#| echo: false #“Run the code, but don’t show the code in the output.”
bigrams_filter6 <- LiteratureNgram |> #LiteratureNgram contains bigram tokenized words
  separate(bigram, c("firstword", "secondword"), sep = " ") |> #separate the bigram column
  filter(!firstword %in% stop_words$word, #remove all the stop-words from the bigram column 
         !secondword %in% stop_words$word) |> #remove all the stop-words from the bigram column 
  filter(!is.na(firstword) & !is.na(secondword)) |> #remove all the missing words 
  count(firstword, secondword, sort = TRUE) |> #count how many times each pair, one in the firstword column, the other from the secondword column. 
  slice_head(n = 200) #keep only 200 rows 

#CREATE GRAPH 
bigram_graph8 <- bigrams_filter6 |>
  filter(n > 1) |>
  graph_from_data_frame() #Converts the data frame into a network graph object

#Makes the graph layout reproducible and creates an arrow style for network edges. 
set.seed(2020)
a <- grid::arrow(type = "closed", length = unit(.15, "inches"))

library(ggtext)

ggraph(bigram_graph8, layout = "fr") +
  geom_edge_link(aes(edge_alpha = n), show.legend = FALSE,
                 arrow = a, end_cap = circle(.07, 'inches')) +
  geom_node_point(color = "blue", size = 5) +
  geom_node_text(aes(label = name), vjust = 1, hjust = 1) +
  labs(
    title = "<strong style='font-size:18px; color:black;'>10 Example Bigram Words Without Stopwords</strong>",
    subtitle = "Connections between words appearing together in the text"
  ) +
  theme_void() +
  theme(
    plot.title = element_markdown(),
    plot.subtitle = element_text(size = 12)
  )
```

This is a directed network graph illustrating common word pairs (bigrams) within the dataset. Each green node represents a unique word, while the arrows indicate the sequence and directional relationship of phrases appearing together in the text. The density of the arrows reveals thematic clusters, most notably a central focus on racial and ethnic disparities and a secondary cluster focused on maternal health and obstetric outcomes. Larger connections represent more frequent word pairings.

### Reference 
* Academic Literature on Racial and Ethnic Disparities in Reproductive Medicine in the US. [see the source here](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-02-25/readme.md)
* Stopwords: [online source](https://juliasilge.github.io/tidytext/reference/get_stopwords.html)
* get_sentiments("bing") R Package
* Sentiment: [online source](https://www.cs.uic.edu/~liub/FBS/sentiment-analysis.html)



